% Chapter 7: Implementation
% Contains: Components, Gossip Protocol, Data Flow Examples

% ============================================================
\section{Implementation Components}
% ============================================================

The three Avena subsystems (Overlay Network, Messaging Network, Fleet Management) are realized through the following software components.

\subsection{avenad (Device Daemon)}

The avenad daemon runs on every Avena device. Configuration is via TOML file (default: \texttt{/etc/avena/avenad.toml}).

\textbf{Default configuration:}

\begin{lstlisting}[language=Rust]
struct AvenadConfig {
    tunnel_mode: TunnelMode,        // default: Kernel
    underlay: UnderlayConfig,       // physical interface filtering
    wireguard: WireguardConfig,     // port allocation
    persistent_keepalive: u16,      // default: 25 seconds
    dead_peer_timeout_secs: u64,    // default: 180 seconds
    keypair_path: Option<PathBuf>,  // optional: persist device identity
    network: NetworkConfig,         // default: fd00:a0e0:a000::/48
    discovery: DiscoveryConfig,     // mDNS enabled by default
    routing: RoutingConfig,         // babeld configuration
    root_cert_path: PathBuf,        // required: trusted CA
    device_cert_path: PathBuf,      // required: device certificate
}

struct UnderlayConfig {
    // Glob patterns for interfaces to include (e.g., "wlan*", "eth*")
    include: Vec<String>,           // default: ["wlan*", "wwan*", "eth*", "enp*"]
    // Glob patterns for interfaces to exclude (takes precedence)
    exclude: Vec<String>,           // default: ["docker*", "veth*", "virbr*", "br-*"]
}

struct WireguardConfig {
    port_range_start: u16,          // default: 51820
    port_range_end: u16,            // default: 52820
    handshake_port_offset: u16,     // default: 1000 (handshake = wg_port + offset)
}

struct RoutingConfig {
    babel_binary: PathBuf,          // default: /usr/sbin/babeld
    babel_socket: PathBuf,          // default: /run/avena/babel.sock
    hello_interval_ms: u32,         // default: 4000
    update_interval_ms: u32,        // default: 16000
}
\end{lstlisting}

\textbf{Interface naming:} WireGuard interfaces are named \texttt{av-<hash8>} where \texttt{<hash8>} is the first 8 characters of a hash derived from \texttt{(local\_device\_id, peer\_device\_id, underlay\_name)}. This keeps names short while ensuring uniqueness.

\textbf{Port allocation:} Each WireGuard interface requires a UDP listen port. Ports are allocated deterministically from the configured range using a hash of the interface name, ensuring consistency across restarts.

The avenad daemon implements all three subsystems:

\textbf{Overlay Network responsibilities:} Underlay Monitoring (netlink events for interface up/down); Peer Discovery (mDNS/DNS-SD per underlay, configured endpoints, LoRa presence); Tunnel Management (certificate exchange, Wireguard key derivation, per-peer-per-underlay interface creation); Routing (babeld in unicast mode); LoRa Channel (presence beacons, gossip updates).

\textbf{Messaging Network responsibilities:} NATS Endpoint (local NATS-compatible API); Interest Management (validation, gossip propagation, tombstones); Message Routing (cost-aware path selection, store-and-forward, deduplication); Geo-Spatial Filtering.

\textbf{Fleet Management responsibilities:} Identity Management (Ed25519 keypair, Device ID); Certificate Management (storage, validation, trust); State Synchronization (CRDT store, gossip protocol); Workload Reconciliation (quadlet configs, container lifecycle); Sidecar Management (spawn avena-sidecar for Addressable workloads, derive keys, push peer updates); Nested Avena Support (derived keypairs for workload\_spawner role).

\subsection{avenactl (CLI Tool)}

Command-line interface for operators: Context Management (connection profiles); Device Operations (list, ping, status, inspect); Workload Management (apply, delete, start/stop, logs); Certificate Management (issue, revoke); Network Diagnostics (links, routes, gossip state, tunnel status).

\subsection{avena-overlay (Library)}

Shared Rust library providing the overlay network foundation:

\begin{lstlisting}[language=Rust]
// Module structure
pub mod identity;     // DeviceId, DeviceKeypair, derived keypairs
pub mod address;      // NetworkConfig, IPv6 address derivation
pub mod crypto;       // Handshake, certificates, key derivation
pub mod discovery;    // mDNS, static peers, DiscoveryService
pub mod tunnel;       // TunnelBackend trait, PeerConfig
pub mod wg;           // WireGuard kernel/userspace implementations
pub mod routing;      // BabeldController, unicast config
pub mod underlay;     // UnderlayManager, netlink monitoring
pub mod daemon;       // AvenadConfig, PeerState

// Key types re-exported at crate root
pub use identity::{DeviceId, DeviceKeypair};
pub use address::NetworkConfig;
pub use crypto::{HandshakeMessage, EphemeralKeypair, CertValidator};
pub use discovery::{DiscoveryService, DiscoveredPeer, Capability};
pub use tunnel::{TunnelBackend, PeerConfig, TunnelError};
pub use routing::BabeldController;
pub use underlay::UnderlayManager;
\end{lstlisting}

\textbf{Dependencies:} \texttt{ed25519-dalek}, \texttt{x25519-dalek}, \texttt{sha2}, \texttt{hkdf}, \texttt{jsonwebtoken}, \texttt{mdns-sd}, \texttt{tokio}, \texttt{netlink-packet-wireguard} (Linux).

\subsection{avena-sidecar (Network Sidecar)}

Lightweight container for Addressable-mode workloads. The host avenad spawns avena-sidecar alongside workloads that require direct overlay IP connectivity.

\textbf{Responsibilities:} Wireguard endpoint management (userspace implementation); identity handling (derived keys from host); peer configuration (receives updates from host avenad); tunnel establishment with remote peers.

\textbf{Non-responsibilities:} Peer discovery (host handles); Babel routing (host handles); gossip participation (host handles). The sidecar is intentionally minimal---it manages only the workload's Wireguard interfaces.

\textbf{Communication with host:} The sidecar connects to the host's NATS endpoint using a JWT issued by avenad. The JWT scopes the sidecar's permissions to its control subject (\texttt{avena.sidecar.<workload-id>.control}). When the host learns of new peers (via gossip), it publishes peer configuration updates to the sidecar's control subject.

% ============================================================
\section{Gossip Protocol}
% ============================================================

\subsection{Synchronization Protocol}

Upon connection between devices $d_1$ and $d_2$:

\textbf{Phase 1 (Authentication):} Exchange certificates $\text{cert}(d_1)$, $\text{cert}(d_2)$. Verify certificate chains to trusted root. Establish Wireguard tunnel; Babel routes converge.

\textbf{Phase 2 (Digest Exchange):} $d_1$ sends state digest: vector of (collection\_id, version\_vector) pairs. $d_2$ sends state digest similarly.

\textbf{Phase 3 (Reconciliation):} For each CRDT collection where digests differ: exchange deltas needed for convergence; verify signatures on received entries; verify issuer certificates authorize the entries; apply CRDT merge operations to local state.

\textbf{Phase 4 (Store-and-Forward):} For each pending application message $m$ with interest $i$ in outbox: determine if peer can deliver $m$ closer to $\text{owner}(i)$ based on relay policies; if so, forward $m$ to peer; peer acknowledges receipt; remove acknowledged messages from local outbox if no other interests require them.

% ============================================================
\section{Data Flow Examples}
% ============================================================

\subsection{Workload Deployment via Mobile Relay}

\begin{Verbatim}[frame=single,fontsize=\small]
Operator          Gateway            Tractor              Sensor
   |                 |                  |                    |
   | apply workload ->                  |                    |
   | selector: sensor                   |                    |
   |                 |                  |                    |
   |                 | store + queue    |                    |
   |                 | (sensor offline) |                    |
   |                 |                  |                    |
   |                 |<-- tractor ------+                    |
   |                 |    connects      |                    |
   |                 |                  |                    |
   |                 | gossip sync ---->|                    |
   |                 | (workload spec)  |                    |
   |                 |                  |                    |
   |                 |    tractor moves to field            |
   |                 |                  |                    |
   |                 |                  |<--- sensor in ---->|
   |                 |                  |     LoRa range     |
   |                 |                  |                    |
   |                 |                  | gossip sync ------>|
   |                 |                  | (workload spec)    |
   |                 |                  |                    |
   |                 |                  |        deploy ---->|
\end{Verbatim}

\subsection{LoRa-Assisted Convergence}

\begin{Verbatim}[frame=single,fontsize=\small]
Sensor (LoRa + WiFi)        Gateway (LoRa + WiFi)
        |                            |
        | presence (LoRa) ---------->|
        | "sensor online,            |
        |  have data for app.soil"   |
        |                            |
        |                            | prepare: 
        |                            | queue workload update
        |                            | for sensor
        |                            |
        |<-------- WiFi establishes --|
        |                            |
        |<-- immediate gossip sync --|
        |    (workload already queued)
        |                            |
        |-- data transfer ---------->|
\end{Verbatim}

\subsection{V2X Emergency Preemption}

\begin{Verbatim}[frame=single,fontsize=\small]
Ambulance           RSU/Signal          Passenger Vehicle
    |                   |                      |
    | publish preempt ->|                      |
    | (geo: within      |                      |
    |  500ft of signal) |                      |
    |                   |                      |
    |                   | adjust timing        |
    |                   |                      |
    |                   | broadcast preempt -->|
    |                   | notification         |
    |                   |                      |
    |                   |              vehicle adjusts
\end{Verbatim}

